Bayesian multivariate skew-normal finite mixture model for analysis of infant development trajectories

### 2019-07-01 manuscript/manuscript_06-30-2019_ca.tex
section 3

%  ALWAYS USE THE referee OPTION WITH PAPERS SUBMITTED TO BIOMETRICS!!!
%  You can see what your paper would look like typeset by removing
%  the referee option.  Because the typeset version will be in two
%  columns, however, some of your equations may be too long. DO NOT
%  use the \longequation option discussed in the user guide!!!  This option
%  is reserved ONLY for equations that are impossible to split across 
%  multiple lines; e.g., a very wide matrix.  Instead, type your equations 
%  so that they stay in one column and are split across several lines, 
%  as are almost all equations in the journal.  Use a recent version of the
%  journal as a guide. 
%  
\documentclass[useAMS,referee]{biom}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{blindtext}
\usepackage{booktabs}
\usepackage{breqn}
\usepackage{algorithm}
\usepackage{algcompatible}
\usepackage{algpseudocode}
\usepackage{float}
\usepackage{dsfont}
\usepackage{dutchcal}
\usepackage{amsfonts}

@@

To make use of P\'olya--Gamma data augmentation, we first write the full conditional distribution of $\boldsymbol\delta_k$ as the prior for $\boldsymbol\delta_k$ times the multinomial likelihood for all $z_i$. $$p(\boldsymbol\delta_k|\mathbf{z},\boldsymbol\delta_{h \ne k}) \propto p(\boldsymbol\delta_k) \prod_{i = 1}^{n} \pi_{ik}^{U_{ik}}(1-\pi_{ik})^{1-U_{ik}},$$
where $p(\boldsymbol\delta_k)$ denotes the prior distribution of $\boldsymbol\delta_k$, $U_{ik} = {1}_{z_i = k}$ is an indicator that subject $i$ belongs to cluster $k$, and $\pi_{ik}$ is defined as in Section 3.4. We can rewrite $\pi_{ik}$ as follows
$$\pi_{ik} = P(U_{ik} = 1) = \frac{\text{e}^{\mathbf{w}_i^T \boldsymbol\delta_k - {c}_{ik}}}{1 + \text{e}^{\mathbf{w}_i^T \boldsymbol\delta_k - {c}_{ik}}} = \frac{\text{e}^{\eta_{ik}}}{1 + \text{e}^{\eta_{ik}}}$$
where ${c}_{ik} = \log \sum_{h \ne k} \text{e}^{\mathbf{w}_i^T \boldsymbol\delta_{h}}$ and $\eta_{ik} = \mathbf{w}_i^T \boldsymbol\delta_k - {c}_{ik}$. We note that the sum $\sum_{h \ne k} \text{e}^{\mathbf{w}_i^T \boldsymbol\delta_{h}}$ includes the reference category, but since we fix $\boldsymbol\delta_K = \mathbf{0}$, we have $\text{e}^{\mathbf{w}_i^T \boldsymbol\delta_K} = 1$, and hence
$$c_{ik} = \log \sum_{h \ne k} \text{e}^{\mathbf{w}_i^T \boldsymbol\delta_{h}} = \log \left ( 1 + \sum_{h \notin \{k,K \}} \text{e}^{\mathbf{w}_i^T \boldsymbol\delta_{h}} \right )$$

We can use the quantities to re-express the full conditionals for $\boldsymbol\delta_k$ as
\begin{eqnarray}
	p(\boldsymbol\delta_k|\mathbf{U}_k,\boldsymbol\delta_{h \ne k}) &\propto& p(\boldsymbol\delta_k) \prod_{i = 1}^{n} \left (\frac{\text{e}^{\eta_{ik}}}{1 + \text{e}^{\eta_{ik}}} \right )^{U_{ik}} \left (\frac{1}{1 + \text{e}^{\eta_{ik}}} \right )^{1-U_{ik}} \nonumber\\ &=& p(\boldsymbol\delta_k) \prod_{i = 1}^n \frac{(\text{e}^{\eta_{ik}})^{U_{ik}}}{1 + \text{e}^{\eta_{ik}}} \label{eq:pglogit}
\end{eqnarray}
which we note is essentially a logistic regression likelihood. We thus apply this P\'olya--Gamma data augmentation scheme from Polson et al. (2013) to update each $\boldsymbol\delta_k$ ($k = 1,...,K-1$) one at a time based on the binary indicators $U_{ik}$.

\subsection{Conditional MSN Imputation}
To accommodate missing at random (MAR) responses, we propose a convenient imputation algorithm that can be implemented ``online'' as part of the Gibbs sampler. In Section 6, we discuss extensions to allow for non-ignorable missingness. Suppose $\mathbf{y}_i$ has $q_i\in (1,\ldots,J)$ observed values, denoted $\mathbf{y}^{obs}_i$, and $J-q_i$ intermittent missing values, denoted $\mathbf{y}^{miss}_i$. We can make use of the stochastic representation given in equation (\ref{eq:msnreg}) to impute $\mathbf{y}^{miss}_i$ from its conditional multivariate normal distribution given $(z_i,t_i,\mathbf{y}^{obs}_i)$:
\begin{eqnarray}
\mathbf{y}^{miss}_i|(z_i=k,t_i,\mathbf{y}^{obs}_i)&\sim& \text{N}_{J-q_i}(\boldsymbol\mu^{cond}_{ki},\boldsymbol\Sigma^{cond}_k),~\text{where}\nonumber\\
\boldsymbol\mu^{cond}_{ki}&=& \boldsymbol\mu^{miss}_{ki} + \boldsymbol\Sigma_{12} \boldsymbol\Sigma_{22}^{-1}(\mathbf{y}^{obs}_i - \boldsymbol\mu^{obs}_{ki}) \label{eq:impute}\\
\boldsymbol\Sigma^{cond}_k&=& \boldsymbol\Sigma_{11} - \boldsymbol\Sigma_{12}\boldsymbol\Sigma_{22}^{-1}\boldsymbol\Sigma_{21},\nonumber \\
\boldsymbol\mu_{ki} &=& \begin{pmatrix} \boldsymbol\mu_{ki}^{miss} \\ \boldsymbol\mu_{ki}^{obs} \end{pmatrix}, \ \text{and} \nonumber \\
\boldsymbol\Sigma_k &=& \begin{pmatrix} 
	\boldsymbol\Sigma_{k11} & \boldsymbol\Sigma_{k12}\\
	\boldsymbol\Sigma_{k21} & \boldsymbol\Sigma_{k22}
\end{pmatrix}, \ \text{where} \nonumber
\end{eqnarray}
$\boldsymbol\Sigma_{k11}$ is a $J-q_i \times J-q_i$ matrix containing the rows and columns of $\boldsymbol\Sigma_k$ corresponding to inicies of $\mathbf{y}_{i}$ where missingness occurs. Similarly, $\boldsymbol\Sigma_{k12}$ is a $J-q_i \times q_i$ matrix containing the rows of $\boldsymbol\Sigma_k$ that correspond to missing indices of $\mathbf{y}_i$, but columns of $\boldsymbol\Sigma_k$ that correspond to observed indices of $\mathbf{y}_i$. The remaining partitions $\boldsymbol\Sigma_{k21}, \ \text{and} \ \boldsymbol\Sigma_{k22}$ are defined in the same manner. Likewise, $\boldsymbol\mu_{ki} = \mathbf{X}_i \boldsymbol\beta_k$ is partitioned into $\boldsymbol\mu_{ki}^{miss}$ and $\boldsymbol\mu_{ki}^{obs}$ with respect to the missing and observed indices of $\mathbf{y}_i$, respectively. These results follow from conventional multivariate normal theory, which we detail in the Web Appendix. An attractive feature of this imputation algorithm is that it avoids multiplicative run-time scaling in $m$, the number of imputations (Gelman \textit{et al.} 2013; Zhou and Reiter, 2010). Our approach also provides more opportunities to explore the missing data parameter space than does multiple imputation, since missing values are drawn at each MCMC iteration, and often in practice $n_{sim} >> m$, where $n_{sim}$ is the number of MCMC iterations (\textbf{find a reference}). In Section 4, we conduct simulation studies to demonstrate that imputing the missing MSN responses improves inferences over available-case analysis.

@@

\paragraph{\textbf{Step 3:}P\'olya--Gamma Data Augmentation for $z_i$} The sampler concludes with updates of the multinomial regression parameters $\boldsymbol\delta_k$, for $k = 1,...,K-1$, followed by updates of each latent cluster indicator $z_i$ $(i=1,\ldots,n)$ from its multinomial logit full conditional. First, we define $U_{ik} = {1}_{z_i = k}$ for $i = 1,...,n$ and $k = 1,...,K-1$. Next, we update $\mathcal{w}_{ik}$ from a $\text{PG}(1,\eta_{ik})$ density, where $\eta_{ik} = \mathbf{w}_i^T \boldsymbol\delta_k - {c}_{ik}$, and ${c}_{ik} = \log \sum_{h \ne k} \text{e}^{\mathbf{w}_i^T \boldsymbol\delta_{h}}$. Next, define $\gamma_{ik} = \frac{u_{ik} - 1/2}{\mathcal{w}_{ik}}$ and let $\boldsymbol\gamma_k = (\gamma_{1k},...,\gamma_{nk})^T$. Finally, for $k = 1,...,K-1$, update $\boldsymbol\delta_k$ from a $\text{N}_r(\mathbf{d}_k,\mathbf{S}_k)$, where $\mathbf{S}_k = (\mathbf{S}_{0k} + \mathbf{W}^T \mathcal{W}_k \mathbf{W})^{-1}$, $\mathcal{W}_k = \text{Diag}(\mathcal{w}_{1k},...,\mathcal{w}_{nk})$,  $\mathbf{d}_k = \mathbf{S}_k (\mathbf{S}_{0k}\mathbf{d}_{0k} + \mathbf{W}^T \mathcal{W}_k \boldsymbol\gamma_k)$, and $\mathbf{W}$ is the $n \times r$ matrix of multinomial logit regression covariates such that the $i^{th}$ row of $\mathbf{W}$ is $\mathbf{w}_i$.

Lastly, we update $z_1,...,z_n$ by first computing $\boldsymbol\pi_i = (\pi_{1i},...,\pi_{Ki})$ as
$$\pi_{ik} = \frac{\text{e}^{\eta_{ik}}}{1 + \sum_{h=1}^{K-1} \text{e}^{\eta_{ih}}},$$ 
where $\eta_{ik} = \mathbf{w}_i\boldsymbol\delta_k$ for $i = 1,...n$ and $k = 1,...,K$. We also compute according to the multivariate Normal density of $\mathbf{y}_i|t_i$ the probability $P(\mathbf{y}_i|\boldsymbol\zeta^*_{ki},\boldsymbol\Sigma_k)$, where $\boldsymbol\zeta_{ki} = \mathbf{x}_i^*\boldsymbol\beta^*_k$. We use these quantities to compute $\boldsymbol\varrho_{i} = (\varrho_{1i},...,\varrho_{Ki})$, where  
$$\varrho_{ik} = P(z_i = k|\mathbf{y}_i,\boldsymbol\zeta^*_i,\boldsymbol\Sigma_k) = \frac{\pi_{ik}P(\mathbf{y}_i|\boldsymbol\zeta^*_{ki},\boldsymbol\Sigma_k)}{\sum_{h = 1}^{K} \pi_{ih}P(\mathbf{y}_i|\boldsymbol\zeta^*_{hi},\boldsymbol\Sigma_h)}$$.
The cluster labels $z_i$ are updated from a $\text{Multinomial}(1,\boldsymbol\varrho_i)$ density for $i = 1,...n$, thus concluding the Gibbs sampler.

>>>>

%  ALWAYS USE THE referee OPTION WITH PAPERS SUBMITTED TO BIOMETRICS!!!
%  You can see what your paper would look like typeset by removing
%  the referee option.  Because the typeset version will be in two
%  columns, however, some of your equations may be too long. DO NOT
%  use the \longequation option discussed in the user guide!!!  This option
%  is reserved ONLY for equations that are impossible to split across 
%  multiple lines; e.g., a very wide matrix.  Instead, type your equations 
%  so that they stay in one column and are split across several lines, 
%  as are almost all equations in the journal.  Use a recent version of the
%  journal as a guide. 
%  
\documentclass[useAMS,referee]{biom}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{blindtext}
\usepackage{booktabs}
\usepackage{breqn}
\usepackage{algorithm}
\usepackage{algcompatible}
\usepackage{algpseudocode}
\usepackage{float}
\usepackage{dutchcal}
\usepackage{amsfonts}

@@

To make use of P\'olya--Gamma data augmentation, we first write the full conditional distribution of $\boldsymbol\delta_k$ as the prior for $\boldsymbol\delta_k$ times the multinomial likelihood for all $z_i$. $$p(\boldsymbol\delta_k|\mathbf{U}_k,\boldsymbol\delta_{h \ne k}) \propto p(\boldsymbol\delta_k) \prod_{i = 1}^{n} \pi_{ki}^{U_{ki}}(1-\pi_{ki})^{1-U_{ki}},$$
where $p(\boldsymbol\delta_k)$ denotes the prior distribution of $\boldsymbol\delta_k$, $\mathbf{U}_k = (U_{k1},...,U_{kn})$, $U_{ik} = {1}_{z_i = k}$ is an indicator that subject $i$ belongs to cluster $k$, and $\pi_{ki}$ is defined as in Section 3.4. We can rewrite $\pi_{ki}$ as follows
$$\pi_{ki} = P(U_{ki} = 1) = \frac{\text{e}^{\mathbf{w}_i^T \boldsymbol\delta_k - {c}_{ki}}}{1 + \text{e}^{\mathbf{w}_i^T \boldsymbol\delta_k - {c}_{ki}}} = \frac{\text{e}^{\eta_{ki}}}{1 + \text{e}^{\eta_{ki}}}$$
where ${c}_{ik} = \log \sum_{h \ne k} \text{e}^{\mathbf{w}_i^T \boldsymbol\delta_{h}}$ and $\eta_{ik} = \mathbf{w}_i^T \boldsymbol\delta_k - {c}_{ki}$. We note that the sum $\sum_{h \ne k} \text{e}^{\mathbf{w}_i^T \boldsymbol\delta_{h}}$ includes the reference category, but since we fix $\boldsymbol\delta_K = \mathbf{0}$, we have $\text{e}^{\mathbf{w}_i^T \boldsymbol\delta_K} = 1$, and hence
$$c_{ki} = \log \sum_{h \ne k} \text{e}^{\mathbf{w}_i^T \boldsymbol\delta_{h}} = \log \left ( 1 + \sum_{h \notin \{k,K \}} \text{e}^{\mathbf{w}_i^T \boldsymbol\delta_{h}} \right )^T$$. 
We can use these quantities to re-express the full conditionals for $\boldsymbol\delta_k$ as
\begin{eqnarray}
	p(\boldsymbol\delta_k|\mathbf{U}_k,\boldsymbol\delta_{h \ne k}) &\propto& p(\boldsymbol\delta_k) \prod_{i = 1}^{n} \left (\frac{\text{e}^{\eta_{ki}}}{1 + \text{e}^{\eta_{ki}}} \right )^{U_{ki}} \left (\frac{1}{1 + \text{e}^{\eta_{ki}}} \right )^{1-U_{ki}} \nonumber\\ &=& p(\boldsymbol\delta_k) \prod_{i = 1}^n \frac{(\text{e}^{\eta_{ki}})^{U_{ki}}}{1 + \text{e}^{\eta_{ki}}} \label{eq:pglogit}
\end{eqnarray}
which is essentially a logistic regression likelihood. We thus apply this P\'olya--Gamma data augmentation scheme from Polson et al. (2013) to update each $\boldsymbol\delta_k$ ($k = 1,...,K-1$) one at a time based on the binary indicators $U_{ki}$. To accomplish this, we define $\mathbf{U}^*_{k} = \left( \frac{U_{k1}-1/2}{w_{k1}},...,\frac{U_{kn}-1/2}{w_{kn}} \right )$. Polson et al. (2013) show that $\mathbf{U}^*_{k}$ follows a multivariate normal distribution with mean $\boldsymbol\eta_k = (\eta_{k1},...,\eta_{kn})$ and covariance matrix $\mathbf{O}_k = \text{Diag}(w_{k1},...,w_{kn})$. Thus, it follows that the full conditional distribution of $\boldsymbol\delta_k$ is given by 
\begin{equation}
	p(\boldsymbol\delta_k|\mathbf{z},\mathbf{W}, \mathbf{O}_k) \propto p_0(\boldsymbol\delta_k) \text{exp} \left [ - \frac{1}{2} (\mathbf{U}^*_k - \mathbf{W}\boldsymbol\delta_k)^T \mathbf{O}_k (\mathbf{U}^*_k - \mathbf{W}\boldsymbol\delta_k)\right].
\end{equation}
By assuming a normal prior for $\boldsymbol\delta_k$, as detailed in Section 3.4, the full conditional distribution for $\boldsymbol\delta_k$ is normal with parameters that follow standard forms.
\subsection{Conditional MSN Imputation}
To accommodate missing at random (MAR) responses, we propose a convenient imputation algorithm that can be implemented ``online'' as part of the Gibbs sampler. In Section 6, we discuss extensions to allow for non-ignorable missingness. Suppose $\mathbf{y}_i$ has $q_i\in (1,\ldots,J)$ observed values, denoted $\mathbf{y}^{obs}_i$, and $J-q_i$ intermittent missing values, denoted $\mathbf{y}^{miss}_i$. We can make use of the stochastic representation given in equation (\ref{eq:msnreg}) to impute $\mathbf{y}^{miss}_i$ from its conditional multivariate normal distribution given $(z_i,t_i,\mathbf{y}^{obs}_i)$:
\begin{eqnarray}
\mathbf{y}^{miss}_i|(z_i=k,t_i,\mathbf{y}^{obs}_i)&\sim& \text{N}_{J-q_i}(\boldsymbol\mu^{cond}_{ki},\boldsymbol\Sigma^{cond}_k),~\text{where}\nonumber\\
\boldsymbol\mu^{cond}_{ki}&=& \boldsymbol\mu^{miss}_{ki} + \boldsymbol\Sigma_{12} \boldsymbol\Sigma_{22}^{-1}(\mathbf{y}^{obs}_i - \boldsymbol\mu^{obs}_{ki}) \label{eq:impute}\\
\boldsymbol\Sigma^{cond}_k&=& \boldsymbol\Sigma_{11} - \boldsymbol\Sigma_{12}\boldsymbol\Sigma_{22}^{-1}\boldsymbol\Sigma_{21},\nonumber \\
\boldsymbol\mu_{ki} &=& \begin{pmatrix} \boldsymbol\mu_{ki}^{miss} \\ \boldsymbol\mu_{ki}^{obs} \end{pmatrix}, \ \text{and} \nonumber \\
\boldsymbol\Sigma_k &=& \begin{pmatrix} 
	\boldsymbol\Sigma_{k11} & \boldsymbol\Sigma_{k12}\\
	\boldsymbol\Sigma_{k21} & \boldsymbol\Sigma_{k22}
\end{pmatrix}, \ \text{where} \nonumber
\end{eqnarray}
$\boldsymbol\Sigma_{k11}$ is a $J-q_i \times J-q_i$ matrix containing the rows and columns of $\boldsymbol\Sigma_k$ corresponding to inicies of $\mathbf{y}_{i}$ where missingness occurs. Similarly, $\boldsymbol\Sigma_{k12}$ is a $J-q_i \times q_i$ matrix containing the rows of $\boldsymbol\Sigma_k$ that correspond to missing indices of $\mathbf{y}_i$, but columns of $\boldsymbol\Sigma_k$ that correspond to observed indices of $\mathbf{y}_i$. The remaining partitions $\boldsymbol\Sigma_{k21}, \ \text{and} \ \boldsymbol\Sigma_{k22}$ are defined in the same manner. Likewise, $\boldsymbol\mu_{ki} = \mathbf{X}_i \boldsymbol\beta_k$ is partitioned into $\boldsymbol\mu_{ki}^{miss}$ and $\boldsymbol\mu_{ki}^{obs}$ with respect to the missing and observed indices of $\mathbf{y}_i$, respectively. These results follow from conventional multivariate normal theory, which we detail in the Web Appendix. An attractive feature of this imputation algorithm is that it avoids multiplicative run-time scaling in $m$, the number of imputations (Gelman \textit{et al.} 2013; Zhou and Reiter, 2010). Our approach also provides more opportunities to explore the missing data parameter space than does multiple imputation, since missing values are drawn at each MCMC iteration, and often in practice $n_{sim} >> m$, where $n_{sim}$ is the number of MCMC iterations (\textbf{find a reference}). In Section 4, we conduct simulation studies to demonstrate that imputing the missing MSN responses improves inferences over available-case analysis.

@@

\paragraph{\textbf{Step 3:}P\'olya--Gamma Data Augmentation for $z_i$} The sampler concludes with updates of the multinomial regression parameters $\boldsymbol\delta_k$, for $k = 1,...,K-1$, followed by updates of each latent cluster indicator $z_i$ $(i=1,\ldots,n)$ from its multinomial logit full conditional. First, we define $U_{ki} = {1}_{z_i = k}$ for $i = 1,...,n$ and $k = 1,...,K-1$. Next, we update ${w}_{ki}$ from a $\text{PG}(1,\eta_{ki})$ density, where $\eta_{ki} = \mathbf{w}_i^T \boldsymbol\delta_k - {c}_{ki}$, and ${c}_{ki} = \log \sum_{h \ne k} \text{e}^{\mathbf{w}_i^T \boldsymbol\delta_{h}}$. Next, define $U^*_{ki} = \frac{U_{ki} - 1/2}{{w}_{ki}}$ and let $\mathbf{U}^*_k = (U^*_{k1},...,U^*_{kn})^T$. Finally, for $k = 1,...,K-1$, update $\boldsymbol\delta_k$ from a $\text{N}_r(\mathbf{d}_k,\mathbf{S}_k)$, where $\mathbf{S}_k = (\mathbf{S}_{k0} + \mathbf{W}^T \mathbf{O}_k \mathbf{W})^{-1}$, $\mathbf{O}_k = \text{Diag}({w}_{k1},...,{w}_{kn})$,  $\mathbf{d}_k = \mathbf{S}_k (\mathbf{S}_{k0}\mathbf{d}_{k0} + \mathbf{W}^T \mathbf{O}_k \boldsymbol\gamma_k)$, and $\mathbf{W}$ is the $n \times r$ matrix of multinomial logit regression covariates such that the $i^{th}$ row of $\mathbf{W}$ is $\mathbf{w}_i$.

Lastly, we update $z_1,...,z_n$ by first computing $\boldsymbol\pi_i = (\pi_{1i},...,\pi_{Ki})$ as
$$\pi_{ik} = \frac{\text{e}^{\eta_{ik}}}{1 + \sum_{h=1}^{K-1} \text{e}^{\eta_{ih}}},$$ 
where $\eta_{ik} = \mathbf{w}_i\boldsymbol\delta_k$ for $i = 1,...n$ and $k = 1,...,K$. We also compute according to the multivariate Normal density of $\mathbf{y}_i|t_i$ the probability $P(\mathbf{y}_i|\boldsymbol\zeta^*_{ki},\boldsymbol\Sigma_k)$, where $\boldsymbol\zeta_{ki} = \mathbf{x}_i^*\boldsymbol\beta^*_k$. We use these quantities to compute $\boldsymbol\upsilon_{i} = (\upsilon_{1i},...,\upsilon_{Ki})$, where  
$$\upsilon_{ik} = P(z_i = k|\mathbf{y}_i,\boldsymbol\zeta^*_i,\boldsymbol\Sigma_k) = \frac{\pi_{ik}P(\mathbf{y}_i|\boldsymbol\zeta^*_{ki},\boldsymbol\Sigma_k)}{\sum_{h = 1}^{K} \pi_{ih}P(\mathbf{y}_i|\boldsymbol\zeta^*_{hi},\boldsymbol\Sigma_h)}$$.
The cluster labels $z_i$ are updated from a $\text{Multinomial}(1,\boldsymbol\varrho_i)$ density for $i = 1,...n$, thus concluding the Gibbs sampler.
